Molecular dynamics simulation of synchronization in driven particles

### 2020-11-06 manuscript/manuscript_ajp_2020.tex
updated MD model

\usepackage{amsmath}  % needed for \tfrac, \bmatrix, etc.
\usepackage{amsfonts} % needed for bold Greek, Fraktur, and blackboard bold
\usepackage{graphicx} % needed for figures

@@

  %What problem did you study and why is it important?
  We study synchronization of interacting particles
  confined to a narrow channel 
  driven by an externally applied force.  
  %What methods did you use?
  Using 
  numerical simulations
  we control the particle interactions,
  external force and particle environment
  in order to mimick experimental studies of driven colloidal particles
  confined by light-fields.
  The molecular dynamics simulations
  model the particle dynamics 
  using overdamped equations of motion suitable for a viscous suspension
  of microscopic particles.
  %What were your principal results?
  We observe
  particle synchronization under a variety of conditions,
  including static and dynamic patterns formed on
  periodic and quasiperiod substrates,
  high density kink propagation.
  We note 
  the limits of applied force that
  cause the dynamics to transition to chaotic behavior.
  %What conclusions have you drawn from these results?
  We demonstrate the transition from
  trapped to sliding dynamics in
  quasiperiodic landscapes differs
  from that of a periodic landscape by [TBD],
  and that kinks of high density propagate
  [more/less ?  ] when [what?].
  We include sample code
  and exercises for students
  that include 
  opportunties
  to reproduce our results and propose
  new numerical experiments.
  With only a few particles in two-dimensions,
  the simulation runs quickly,
  making this an appropriate model for undergraduates to explore.

@@

%----------------------------------------------------------------------------
%paper outline
In the following paper we describe
our molecular dynamics model in Section~\ref{sec:MD}.
%including various  and confining environments.
We include code to simulate
and visualize the dynamics in this section
and supplementary material.
In Section~\ref{sec:results} we summarize
our results,
including synchronized motion of a single confined particle
driven across a periodic landscape in 
Section~\ref{sec:one} and
multiple interacting particles
in Section~\ref{sec:sync}.
%we
%demonstrate how uniform environments and applied forces
%and create synchronized flow patterns.
We present these results using standard tools of non-linear oscillators
such phase diagrams of particle velocity versus position.
In Section~\ref{sec:quasiperiod},
we show how an aperiodic landscape modifies the particle dynamics.
Finally we explore the transition to chaotic dynamics in 
in Section~\ref{sec:chaos}.
Our conclusions
are in Section~\ref{sec:conclusion}. 
In each section
we suggest exercises for interested students,
and summarize our suggestions in Section~\ref{sec:problems}.

@@

Particles are subject an external time-dependent driving force
$\vec{F}_{D}(t)$
applied parallel to the y-direction.
We model this force as
\begin{equation}
\vec{F}_{D}(t) = [F_{DC} + F_{AC} \sin(\omega t)] \hat{y},
\end{equation}
with modifiable parameters including
a constant component $F_{DC}$,
and a time dependent component with amplitude $F_{AC}$
and frequency $\omega = 2 \pi f$.
The frequency is scaled in time units,
where [finish].
 We demonstrate how a single particle (Sec.~\ref{sec:one})
 and many particles
 move (Sec.~\ref{sec:twenty})
 in response to this applied force in a variety of light fields.
 In Sec.~\ref{sec:kink} we set $F_{DC}$
 to zero and track the motion
 of a high density area of a particle chain
 (i.e. kink dynamics).
 %The driving force does add energy into the system, and some of it is lost.

 We create several model light fields,
 creating a landscape of potential minima and maxima
 that modify the local force on a particle as a function of position
 $\vec{F}_{l,i} = \vec{F}_{l}(\vec{r}_i).$
 The landscape potential $V(\vec{r}) $ are static,
 generating a conservative force $\vec{F}_l= -\nabla V_l(\vec{r})$
 with fixed minima and maxima
 that are periodic or quasi-periodic,
 as described in Sec.~\ref{sec:results}. 
 One example of a periodic light field is
 %that used in Sec.~\ref{sec:one}
 where 
 \begin{equation}
   V_l = V_0 \cos{(N_p \pi y / L)}
 \end{equation}
 where $N_p$ are the number of troughs in the landscape,
 and $V_0$ is an adjustable parameter
 to set the depth of the troughs.

 In multi-particle simulations, we confine
 the particles along the x-direction with a
 quasi-one-dimensional geometry
 using a periodic function.
 \begin{equation}
   V_{q1D} = V_0 \cos{(\pi x / L)}
 \end{equation}.
 %%We could also use a Gaussian function.

@@

We model the 
particle dynamics 
with an overdamped
equation of motion
integrated with the Verlet method.
Here 
the suspending fluid is highly viscous
and exerts a damping force on the particles equivalent
to a linear drag force.
%Overdamped dynamics is a common assumption in models of colloidal particles.
%This is an assumption that 
The damping comes from a viscous fluid model
providing a nonconservative force,
modeled as a linear friction $F_{drag} = -\eta \vec{v}_i$
sufficient so that
the acceleration of the particle is zero.
[This model is encountered in Ch. 2 of Taylor's Classical Mechanics....]

A single particle has the equation of motion 
\begin{equation}
\eta \vec{v}_i = \vec{F}_{l,i} + \sum_{i \neq j}^{N} \vec{F}_{ij} + \vec{F}_{D}(t).
\end{equation}
where $\eta = 1$.
The equation of motion provides a direct calculation of the velocity
of an individual particle from its location.
Since we model the acceleration $\vec{a}$ as zero,
the Verlet method simplifies to 
the Euler method,
which is used to calculate the position. %Most undergraduate students learn the Euler method in Computational Physics and Differential Equations, but don't encounter the Verlet method.
\begin{equation}
\vec{r}_i(t+\Delta t) = \vec{v}_i(t) \Delta t + \vec{r}_i(t).
\end{equation}

@@

We place a single particle in a periodic landscape 
\begin{equation}
  \label{eq:landscape}
  V_l = V_{0x} \cos{(\pi x/L)} + V_{0y} \cos{(N_p \pi y / L)} 
\end{equation}
where $N_p=20$ are the number of troughs in the landscape,
and $V_{0x}=2$ and $V_{0x}=2$.
This is illustrated in Fig.~\ref{fig:1}(a) where
the red regions show local maxima and the blue local minima.
The code for generating
a two dimensional colored plot
of the potential landscape
is calculated by evaluating
the analytic function in Eq.~\ref{eq:landscape}
for a grid of values $(x_n,y_n)$.
%is included in the supplementary material,

We apply an external applied force, 
with 
a constant $F_{AC}$ with frequency $\omega$ then slowly increase $F_{DC}$.
The particle hops between troughs in the energy landscape
when the superposition of $F_{AC}$ and $F_{DC}$ is large enough
to overcome the barrier height.
By modifying $F_{DC}$ we 
achieve a variety of oscillation modes.
A mode is a periodic pattern of hops
with a constant average particle velocity, $\bar{v}_{y}$
over a range of driving forces $F_{DC}$.
We illustrate mode-locking in 
the velocity-force plot in Fig.~\ref{fig:1}(b).
Here $\bar{v}_{y}$ is increasing in non-uniform steps.
Our simulations reproduce results presented in 
Juniper {\it et al.} \cite{juniper2015}
which demonstrated
mode locking in
experiments of 
driven colloids on a
optical periodic landscape.

We show the dynamics 
in the attached movie, Figure1.mp4.

@@

We simulated a twenty particle system confined to a narrow channel, as shown in Fig. 2a). The interparticle forces of neighboring particles cause the system to form a buckled chain when the system is annealed. When a single particle is driven, the neighboring particles act similarly to a periodic landscape to impede its motion. A driven particle can exhibit mode locking with a well-chosen AC drive and frequency. In the attached movie, Figure2.mp4, we show the complex dynamics of mode locking, where the driven particle leap-frogs past the other particles.

@@

\subsection{Kinked system}
\label{sec:static}	% You can label sections for reference
We confine $N$ particles to $N-1$ troughs to create a
local high density region.
$F_{DC}/F_{AC} = 1$ [CHECK!]

@@

@@

>>>>

\usepackage{amsmath}  % needed for \tfrac, \bmatrix, etc.
\usepackage{amsfonts} % needed for bold Greek, Fraktur, and blackboard bold
\usepackage{graphicx} % needed for figures
\usepackage{pythontex}

@@

  %What problem did you study and why is it important?
  We study synchronization of interacting particles
  confined to a narrow channel 
  driven by an externally applied force.  
  %What methods did you use?
  Using 
  numerical simulations
  we control the particle interactions,
  external force and particle environment
  in order to mimick experimental studies of driven colloidal particles
  confined by light-fields.
  The molecular dynamics simulations
  model the particle dynamics 
  using overdamped equations of motion suitable for a viscous suspension
  of microscopic particles.
  %What were your principal results?
  We observe
  particle synchronization under a variety of conditions,
  including static and dynamic patterns formed on
  periodic and quasiperiod substrates
  and the propagation of 
  high density kinks.
  %What conclusions have you drawn from these results?
  We demonstrate the transition from
  trapped to sliding dynamics in
  quasiperiodic landscapes differs
  from that of a periodic landscape by [TBD],
  and that kinks of high density propagate
  [more/less ?  ] when [what?].
  We also explore 
  the limits of applied force that
  cause the dynamics to transition to chaotic behavior.
  We include sample code
  and exercises for students
  that include 
  opportunties
  to reproduce our results and propose
  new numerical experiments.
  With only a few particles in two-dimensions,
  the simulation runs quickly,
  making this an appropriate model for undergraduates to explore.

@@

%----------------------------------------------------------------------------
%paper outline
In the following paper we describe
our molecular dynamics model in Section~\ref{sec:MD}.
%including various  and confining environments.
We include code to simulate
and visualize the dynamics in this section
and supplementary material.
In Section~\ref{sec:results} we summarize
our results,
including synchronized motion of a single confined particle
driven across a periodic landscape in 
Section~\ref{sec:one} and
multiple interacting particles
in Section~\ref{sec:sync},
including stationary propagation of high density kinks
in Sec.~\ref{sec:kink}.
%we
%demonstrate how uniform environments and applied forces
%and create synchronized flow patterns.
We present these results using standard tools of non-linear oscillators
such phase diagrams of particle velocity versus position.
In Section~\ref{sec:quasiperiod},
we show how an aperiodic landscape modifies the particle dynamics.
Finally we explore the transition to chaotic dynamics in 
Section~\ref{sec:chaos}
and conclude 
in Section~\ref{sec:conclusion}. 
In each section
we suggest exercises for interested students,
and summarize our suggestions in Section~\ref{sec:problems}.

@@

Particles are subject an external time-dependent driving force
$\vec{F}_{D}(t)$
applied parallel to the y-direction.
We model this force as
\begin{equation}
\vec{F}_{D}(t) = [F_{DC} + F_{AC} \sin(\omega t)] \hat{y},
\end{equation}
with modifiable parameters including
a constant component $F_{DC}$,
and a time dependent component with amplitude $F_{AC}$
and frequency $\omega = 2 \pi f$.
The frequency is scaled in time units,
where [finish].
 We demonstrate how a single particle (Sec.~\ref{sec:one})
 and many particles
 move (Sec.~\ref{sec:sync})
 in response to this applied force in a variety of light fields.
 In Sec.~\ref{sec:kink} we set $F_{DC}$
 to zero and track the motion
 of a high density area of a particle chain
 (i.e. kink dynamics).
 %The driving force does add energy into the system, and some of it is lost.

 We create several model light fields,
 creating a landscape of potential minima and maxima
 that modify the local force on a particle as a function of position
 $\vec{F}_{l,i} = \vec{F}_{l}(\vec{r}_i).$
 The landscape potential $V(\vec{r}) $ are static,
 generating a conservative force $\vec{F}_l= -\nabla V_l(\vec{r})$
 with fixed minima and maxima
 that are periodic or quasi-periodic,
 as described in Sec.~\ref{sec:results}. 
 One example of a periodic light field is
 %that used in Sec.~\ref{sec:one}
 where 
 \begin{equation}
   V_l(y) = V_0 \cos{(N_p \pi y / L)}
 \end{equation}
 where $N_p$ are the number of troughs in the landscape,
 and $V_0$ is an adjustable parameter
 to set the depth of the troughs.

 In multi-particle simulations, we confine
 the particles along the $x-$direction with a
 quasi-one-dimensional geometry
 using a periodic function.
 \begin{equation}
   V_{q1D}(x) = V_0 \cos{(\pi x / L)}
 \end{equation}.
 %%We could also use a Gaussian function.

@@

The simulation is controlled by a $for()$ loop
which runs from an intial to maximum $time$ integer.
Each time step represents a simulation time of $\Delta t=0.001$.
At each timestep
we evaluate the net force on each particle as a function of its position
$\vec{r}_i(t)$
and then use the equation of motion to move particles
to an updated position
$\vec{r}_i(t+\Delta t)$.
%
We model the 
particle dynamics 
with an overdamped
equation of motion
integrated with the Verlet method.
Here 
the suspending fluid is highly viscous
and exerts a damping force on the particles equivalent
to a linear drag force.
%Overdamped dynamics is a common assumption in models of colloidal particles.
%This is an assumption that 
The damping comes from a viscous fluid model
providing a nonconservative force,
modeled as a linear friction $\vec{F}_{drag} = -\eta \vec{v}_i$
sufficient so that
the acceleration of the particle is zero.
Such a model is appropriate when the particles
are small and the viscosity is high [finish... reference...]
%[This model is encountered in Ch. 2 of Taylor's Classical Mechanics....]

A single particle has the equation of motion 
\begin{equation}
\eta \vec{v}_i = \vec{F}_{l,i} + \sum_{i \neq j}^{N} \vec{F}_{ij} + \vec{F}_{D}(t).
\end{equation}
where $\eta = 1$.
The equation of motion provides a direct calculation of the velocity
of an individual particle from its location.
Since we model the acceleration $\vec{a}$ as zero,
the Verlet method simplifies to 
the Euler method,
which is used to calculate the position at subsequent time steps.
%Most undergraduate students learn the Euler method in Computational Physics and Differential Equations, but don't encounter the Verlet method.
\begin{equation}
\vec{r}_i(t+\Delta t) = \vec{v}_i(t) \Delta t + \vec{r}_i(t).
\end{equation}

@@

We place a single particle in a periodic landscape with
corrugations along the y-direction 
\begin{equation}
  \label{eq:landscape}
  V_l(y) = V_{0y} \cos{(N_p \pi y / L)} 
\end{equation}
where $N_p=20$ are the number of troughs in the landscape,
and $V_{0x}=2$ and $V_{0x}=2$.
This is illustrated in Fig.~\ref{fig:1}(a) where
the red regions show local maxima and the blue local minima.
The code for generating
a two dimensional colored plot
of the potential landscape
is calculated by evaluating
the analytic function in Eq.~\ref{eq:landscape}
for a grid of values $(x_n,y_n)$.
%is included in the supplementary material,

We apply an external applied force, 
with 
a constant $F_{AC}$ with frequency $\omega = 2\pi f$
then slowly increase $F_{DC}$.
The particle hops between troughs in the energy landscape
when the superposition of $F_{AC}$ and $F_{DC}$ is large enough
to overcome the barrier height.
By modifying $F_{DC}$ we 
achieve a variety of oscillation modes.
A mode is a periodic pattern of hops
with a constant average particle velocity, $\bar{v}_{y}$
over a range of driving forces $F_{DC}$.
We illustrate mode-locking in 
the velocity-force plot in Fig.~\ref{fig:1}(b).
Here $\bar{v}_{y}$ is increasing in non-uniform steps,
with a quantized height of
$\bar{v}_{y} = n \lambda f$,
where $n$ is an integer,
$\lambda = S_Y/N_p = 36.5/20 = 1.825$ is the spatial period, or wavelength
of the landscape,
and $f = 0.01$ cycles per time unit.

Our simulations reproduce results presented in 
Juniper {\it et al.} \cite{juniper2015}
which demonstrated
mode locking in
experiments of 
driven colloids on a
optical periodic landscape.

We illustrate the hopping pattern in Fig.~\ref{fig:1}(c)
and 
show the dynamics 
in supplementary materials \cite{supp1}.

@@

We simulated a twenty particle system confined to a narrow channel, as shown in Fig. 2a).  We create the confining channel with a sinusoidal function
with a single period.
\begin{equation}
  \label{eq:channel}
  V_l(x) = V_{0x} \cos{(\pi x/L)}
\end{equation}
where the trough heights is larger  $V_{0x}$,
and the associated force
\begin{equation}
\vec{F}=-\nabla V(x) = -\frac{dV}{dx} \hat{x} = - \frac{V_{0x} \pi}{L} \sin{(\pi x/L)} \hat{x}
\end{equation}
restores particles to the center of a long narrow region of the simulation.
This is shown in Fig.~\ref{fig:2}(a)
where red regions are high potential
and blue regions are low potential.

The initial configuration of the system is shown i
Fig.~\ref{fig:2}(a).  
We annealed the system into a ground-state configuration
by raising the system to a high temperature $T$,
and slowly lowering the temperature in steps of $dT=-0.01$
until the particles form a buckled chain due to the
competition between particle repulsion and channel confinement.

The interparticle forces of neighboring particles cause the system to form a buckled chain when the system is annealed.

When a single particle is driven, the neighboring particles act similarly to a periodic landscape to impede its motion. A driven particle can exhibit mode locking with a well-chosen AC drive and frequency. In the attached movie, Figure2.mp4, we show the complex dynamics of mode locking, where the driven particle leap-frogs past the other particles.

@@

\subsection{Kinked system}
\label{sec:kink}	% You can label sections for reference
We confine $N$ particles to $N-1$ troughs to create a
local high density region.
$F_{DC}/F_{AC} = 1$ [CHECK!]

@@

\bibitem{supp1} See Figure1.mp4 in appropriate %\url{}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Supplementary Materials}

\subsection{Gridded Contour Plot of landscape}
%code is located in
%~/pymodules/animation_code/channel_colloid_movie_maker.py

%\begin{pyverbatim}
##########################################################
#ADD CONTOUR PLOT
##########################################################
def add_contour(ax,L,N,corrugated = True):
    '''
    Hardwired to color in the quasi1D potential to contain 
    the particles in a trough.  
    Can also add the washboard/corrugated substrate.

    Required Arguments

    Optional Arguments:

    corrugated (default = True)  
    Adds the washboard in the y-direction.  
    Hardwired for a single parameter set.    
    '''

    a_p = L/N

    #assuming Tiare's trough system, so we won't want to cover the entire range
    X = np.arange(0, L/2.0, 0.1)
    Y = np.arange(0, L, 0.1)
    X, Y = np.meshgrid(X, Y)

    Z_mag = 2.0 # set by what "looks good"
    Z = Z_mag*np.sin(2*np.pi*X/L)
    if corrugated == True:
        Z += np.sin(2*np.pi*(Y+1.75)/a_p)

    cmap=cm.coolwarm_r

    #alphs is the degree of transparency, again, set by what looks good.
    cset = ax.contourf(X, Y, Z, cmap=cmap,alpha=0.25)

    #ax1.set_xlim(15,20)
    #ax1.set_ylim(15,20)

    #ax1.set_xlabel(r"$X$")
    #ax1.set_ylabel(r"$y$",rotation='horizontal',ha='right')

    #ax1.set_xticks([])
    #ax1.set_yticks([])
    return
 %   \end{pyverbatim}
